Allosteric modulation by the fatty acid site in the glycosylated SARS-CoV-2 spike

Version 1

Abstract

Abstract

The trimeric spike protein plays an essential role in the SARS-CoV-2 virus lifecycle, facilitating virus entry through binding to the cellular receptor angiotensin-converting enzyme 2 (ACE2) and mediating viral and host membrane fusion. The SARS-CoV-2 spike contains an allosteric fatty acid (FA) binding site at the interface between two neighbouring receptor-binding domains. This site, also found in some other coronaviruses, binds free fatty acids such as linoleic and oleic acid, and other small molecules. Understanding allostery and how this site modulates the behaviour of different regions in this protein could potentiate the development of promising alternative strategies for new coronavirus therapies. Here, we apply dynamical nonequilibrium molecular dynamics (D-NEMD) simulations to investigate allosteric effects and identify the communication pathways in the fully glycosylated spike in the original SARS-CoV-2 ancestral variant. The results reveal the allosteric networks that connect the FA site to important functional regions of the protein, including some more than 40 Å away. These regions include the receptor binding motif, an antigenic supersite in the N-terminal domain, the furin cleavage site, the regions surrounding the fusion peptide and a second allosteric site known to bind heme and biliverdin. The networks identified here highlight the complexity of the allosteric modulation in this protein and reveal a striking and unexpected connection between different allosteric sites. Notably, 65% of amino acid substitutions, deletions and insertions in the Alpha, Beta, Delta, Gamma and Omicron variants map onto or close to the identified allosteric pathways.

Main text

The SARS-CoV-2 virus, like other β-coronaviruses, uses the spike protein to mediate virus entry into host cells. The spike is a trimeric glycoprotein embedded in the virus envelope. During the initial stage of the SARS-CoV-2 infection process, the spike binds to host target cells, primarily via the receptor angiotensin-converting enzyme 2 (ACE2) (1, 2), but it can also bind to other targets, such as neuropilin-1 (3, 4), estrogen receptor α (5) and potentially to nicotinic acetylcholine receptors (6–8) and sugar receptors (9). Given its crucial role in the infection process and the fact that it is one of the main targets for virus neutralisation, the spike has emerged as one of the most important targets for developing COVID-19 therapies and vaccines (e.g. (10–14)).

Each monomer of the SARS-CoV-2 spike is composed of two subunits (S1 and S2), which contain the structural motifs that directly bind to the host receptors as well as those needed for the membrane fusion process (15–17). The spike also contains three fatty acid (FA) binding sites at the interfaces between neighbouring receptor binding domains (RBDs) (18) (Figure 1A). Each FA binding site is defined by a hydrophobic pocket formed by two RBDs, with one RBD providing the aromatic and hydrophobic residues to accommodate the FA hydrocarbon tail and the other providing the polar and positively charged residues that bind the FA carboxylate headgroup (Figure 1B). The essential FA linoleic acid (LA) binds (as linoleate) with high affinity in the FA pocket, stabilising the spike in a locked conformation, in which the RBDs are all ‘down’ with the receptor-binding motifs (RBMs) occluded inside the trimer, and thus inaccessible for binding to ACE2 (18). The discovery of this allosteric site inspired the development of new spike-based therapies based on FAs or other natural, repurposed, or specifically designed small molecules able to bind to the FA site (10, 19–21). Several cryo-EM structures of the SARS-CoV-2 spike in complex with small molecules, such as linoleic, oleic, and all-trans retinoic acid and SPC-14, bound to the FA site, are now available (10, 18–20). Following this discovery, equivalent FA sites have been identified in several closely related coronavirus spikes (18, 22, 23). Surface plasmon resonance experiments and cryo-EM structures show that the FA site is conserved in the spike proteins of highly pathogenic β-coronaviruses, such as SARS-CoV, MERS-CoV, SARS-CoV-2, but not in the spikes of common, mild disease-causing β-coronaviruses (24).

Structure of the glycosylated head region of the ancestral SARS-CoV-2 spike with linoleate bound to the free fatty acid (FA) binding site.

(A) Model of the ectodomain of the glycosylated SARS-CoV-2 spike with linoleate (LA) bound. The spike-LA complex model was built using the cryo-EM structure 7JJI as a reference (23). Each monomer in the spike homotrimer is shown in a different colour: dark blue, light blue and green. The glycans are indicated with grey sticks, and LA molecules are highlighted with magenta spheres. Three FA binding sites exist in the trimer, each located at the interface between two neighbouring monomers. Note that in this model, all three receptor-binding motifs (RBMs) are in the ‘down’ conformation, and the protein is cleaved at the furin recognition site at the S1/S2 interface. (B) Detailed view of the FA binding site. This hydrophobic site is formed by two RBDs, with one providing the hydrophobic pocket for the FA hydrocarbon tail and the other providing polar (Q409) and positively charged (R408 and K417) residues to bind the negatively charged FA headgroup.

Simulations using the dynamical-nonequilibrium molecular dynamics (D-NEMD) approach showed that the FA site allosterically modulates the behaviour of functional motifs in both the ancestral (also known as wild type, ‘early 2020’, or original) spike and in several variants (25–27). These D-NEMD simulations showed that the FA site is allosterically connected to the RBM, N-terminal domain (NTD), furin cleavage site, and the region sourrounding the fusion peptide (FP) (25–27). These regions have significantly different allosteric behaviours when comparing the ancestral, Alpha, Delta, Delta plus, and Omicron BA.1 variants (25–27). They differ not only in the amplitude of the structural response of specific regions in the S1 and S2 subunits to LA removal but also in the rates at which the structural changes propagate (25–27). Some mutations (such as L18F, T19R, G142D, E156Δ–F157Δ, T478K and P681H/R) were shown to enhance connections between important functional motifs and the FA site, but others (such as H69Δ–V70Δ, Y144Δ, and D614G) reduce them (25, 26). Nonetheless, all such previous D-NEMD simulations did not incorporate the spike’s many N- and O- linked glycans. However, glycans are critical to the function of the spike, not only in protecting it from immune recognition, but also in modulating its dynamics (28–31). A crucial unresolved question, therefore, is whether and how glycosylation affects allosteric communication with the FA site within the spike protein.

The spike is heavily glycosylated with 22 N-linked glycosylation sites per monomer (15, 32), of which 18 are constantly modified (32–35). It also contains at least two O-glycosylation sites per monomer with low occupancy (34, 36, 37). This thick glycan coating plays a crucial role in shielding (30, 31) and infection (28–31). For example, the dense layer of weakly immunogenic complex glycans forms a shield around the spike which facilitates evasion of the host immune system (30). Glycans also impact the dynamics and stability of the essential regions of the protein, including the RBDs, and modulate binding to ACE2 (e.g. (28–30)). Here, we use D-NEMD simulations (38, 39) to characterise the response to LA removal of the fully glycosylated SARS-CoV-2 ancestral spike to investigate allosteric modulation by the FA site and the effects of glycans on allosteric behaviour. D-NEMD simulations (38, 39) have proved to be an effective approach to investigate responses of receptors (40–42), identify allosteric effects (25–27, 43–46) and study effects of pH changes (47) in biomolecular systems, including SARS-CoV-2 targets (25–27, 45, 47).

We performed extensive equilibrium MD simulations, followed by hundreds of D-NEMD simulations, to analyse the response of the fully glycosylated, cleaved (at the furin recognition site) spike to LA removal. The model of the locked state (with all NDBs down) with LA bound remained stable over the equilibrium simulations (Figure S1A), showing structural convergence after ∼50 ns and minimal secondary structure loss after 750 ns (Figure S1B). Principal component analysis was performed to check the equilibration and sampling of the replicates (6–8) (Figure S1C). All LA molecules remained stably bound during the equilibrium simulations (Figures S2A and S2D), with the carboxylate head group of LA making consistent salt-bridge interactions with K417 and occasional interactions with R408 (Figures S2B and S2C).

The dynamic behaviour of the glycans over the equilibrium trajectories was also analysed (Figures S3-S4). As previously observed (30), the glycans are very flexible, exhibiting diverse levels of motion depending on their sequence, branching and solvent exposure (Figure S3). Generally, the N-glycans located in the NTD show higher fluctuations than those of the RBD (Figure S3). The glycan linked to N331 from chain C is an exception, with one of the largest RMSF values. The O-glycans connected to T323 and S325, close to the RBD, are less flexible than the N-glycans (Figure S3). The highly dynamic profile of the glycans (Figure S4) helps the spike to evade the host immune response by masking immunogenic epitopes, thus preventing them from being targeted by the host’s neutralising antibodies. To quantify the shielding effect, the spike accessible surface area (ASA) covered by the glycans was determined for probe radii ranging from 0.14 (approximate radius of a water molecule) to 1.1 (approximate radius of a small antibody molecule) nm. As can be seen in Figures S4B and S4C, consistently with previous findings reported by Casalino et al. (30) for the closed state (with all NBDs in the ‘down’ conformation), the spike head has a thick glycan shield, which covers ∼60% of the protein accessible area for a 1.0-nm-radius probe and restricts the binding of medium size molecules to the protein. However, small molecules (probes with a radius 0.14-0.3nm) can potentially penetrate the shield more easily as it only covers ∼26% of the area of the protein accessible to smaller probes (Figure S4C).

An ensemble of 210 conformations (70 configurations per replicate) was extracted from the equilibrium MD simulations and used as starting points for the D-NEMD simulations, which investigated the effect of LA removal (Figure S5). The perturbation used here, the instantaneous removal of LA from the FA sites, is the same as in our previous work (25–27). This perturbation is designed to force the system out of equilibrium, creating the driving force necessary for the conformational changes of interest to occur. LA removal from the FA sites triggers the structural response of the protein as it adapts to an empty FA site. Analysis of the evolution of the structural changes reveals the mechanical and dynamical coupling between the structural elements involved in response to LA removal and identifies the allosteric pathways connecting the FA site to the rest of the protein. The evolution of the structural response of the protein is extracted using the Kubo-Onsager relation (38, 39, 48, 49) from the difference between the equilibrium and nonequilibrium trajectories at equivalent points in time (Figure S5). The response obtained for each pair of equilibrium and nonequilibrium simulations is then averaged over all 210 trajectories, hence reducing noise (38, 39), and allowing the statistical significance of the responses to be assessed from the standard error of the mean (Figure S6-S8) (38).

LA removal initiates a complex chain of structural changes that are, over time, propagated within the protein. The deletion of the LA molecules immediately triggers a structural change in the FA site, which contracts as the sidechains of the residues lining it move closer to each other, filling the space once occupied by the LA molecule (Figure S9). The changes in the FA site are then swiftly transmitted to well defined regions of the protein, notably NTD, RBM and FP-surrounding regions (Figures 2 and S10-S11).

Structural response of the glycosylated spike to LA removal.

The average Cα displacements 0.1, 1 and 10 ns after LA removal from the FA binding sites are shown, mapped onto the starting structure for the equilibrium simulations. The norm of the average Cα displacement vector between the D-NEMD apo and equilibrium LA-bound simulations was calculated for each residue using the Kubo-Onsager relation (38, 39, 48, 49). The final displacement values are the averages obtained over the 210 pairs of simulations (Figures S6-S8). The cartoon thickness and structure colours (scale on the right) indicate the average Cα-positional displacement. Each RBD, NTD, furin site and FP are subscripted with their chain ID (A, B or C). Glycans are shown as light grey sticks, whereas the dark grey spheres highlight the position of the LA molecule. The FA site shown in this figure corresponds to FA site 1, which is located at the interface between chains C and A (see Figures S10-S11 for the responses of the other two FA sites, which are generally similar).

Despite differences in the amplitude of the structural responses, the cascade of events observed here for the fully glycosylated ancestral spike is similar to that of the non-glycosylated protein (25–27) (Figure S12). Figure S12 shows a strong positive correlation between the responses obtained for the non-glycosylated and glycosylated spikes, with the largest differences in the responses between our current and previous D-NEMD simulations being around the furin recognition site. This is expected, because this site is cleaved in the glycosylated simulations and uncleaved in the non-glycosylated ones. The furin cleavage/recognition site is a polybasic four-residue insertion located on a solvent-exposed loop at the S1/S2 junction (15, 16). This site is important for the activation of the spike (50), and its presence affects viral infectivity (e.g. (26, 50–53)).

The evolution of the response of the spike to LA removal reveals the pathways through which structural changes propagate from the FA site to functional motifs (e.g. motifs involved in membrane fusion and antigenic epitopes (Figures 3 and S13-S14)). The structural changes induced by LA removal, which start in the FA site (mainly in the P337-A348 and S366-A372 regions of one monomer and T415-K417 of the other one), are rapidly transmitted to the rest of the RBD. The R454-K458 region is particularly important because it mediates the transmission of the structural changes to the A475-C488 segment in the RBM.

Structural responses of functional regions of the spike.

Close-up view of the structural response of the RBD (A), NTD (B) and FP surrounding regions (C) to LA removal. The FA site shown here is FA site 1, located at the interface between chains C and A (see Supplementary Figures S13-S14 for the responses of the other two FA sites, which are similar). Structure colours and cartoon thickness indicate the average Cα displacement values. Each region is subscripted with its chain ID (A, B or C). The dark grey spheres highlight the FA binding site. In the images representing the spike at t=0 ns (left side images A, B and C), the glycans are shown as light grey sticks, whereas the dark grey spheres highlight the FA binding site. Glycans are omitted from the panels showing the responses at t=0.1, 1 and 10 ns, but are present in the simulations. For more detail, see the legend of Figure 2.

In addition to the amplitude of the structural changes induced by LA removal, the average direction of the motion can also be computed by determining the average displacement vector of Cα atoms (45) between the equilibrium and nonequilibrium trajectories at equivalent time points (Figures 4 and S16-S19). Indeed, the amplitude of the structural changes in Figures 2–3, S10-S11 and S13-S14 corresponds to the norm of the average Cα displacement vector. Upon LA removal, the spike regions that form the FA site show responses with well-defined directions. These segments include P337-A348 and S366-A372, regions that contain residues whose side chains form the FA pocket (Figures 4 and S16-S19). Soon after LA deletion, P337-A348 and S366-A372 move inwards towards the FA site. These motions collectively reflect the contraction of the FA site upon LA removal. The directions of RBM motions are more diverse, with two of the RBMs (namely RBMB and RBMC) displaying an upward movement and the third showing an opposite downward one (RBMA) at t=10 ns (Figures 4 and S16-S19).

Direction of the structural responses of the RBD and FP-surrounding regions to LA removal.

The average Cα displacement vectors at t=10 ns are shown. These vectors were determined by averaging Cα displacement vectors between the equilibrium and nonequilibrium trajectories over the 210 replicas. Vectors with a length ≥0.1 nm are displayed as blue arrows with a scale-up factor of 10. The average displacement magnitudes are represented on a white-yellow-orange-red scale. The dark grey spheres represent the FA site. This figure shows the direction of the responses around FA site 1, which is located at the interface between chains C and A (see Supplementary Figures S18-S19 for the direction of the motions in the other two FA sites).

SARS-CoV-2 spike has two glycans located on the RBD, notably N-glycans at position N331 and N343 (32). The N343 glycan is particularly interesting because it is situated immediately after one of the regions that transmits structural changes from the FA site, namely P337-F342. As well as being close to the FA site, this glycan can also directly bridge the two neighbouring RBDs (Figure S15), which may strengthen the connection between the FA site and these regions. N343 has been shown to play a role in the RBD opening mechanism by acting as a gate for the change from the ‘down’ to the ‘up’ conformation (29).

The structural changes induced by LA removal are also swiftly propagated to the NTD via P337-F342, W353-I358, and C161-P172 (Figures 3 and S13-S14). The response of the protein, which starts in the P337-A348 segment in the FA site, is transmitted to W353-I358, C161-P172 and then to several antigenic epitopes located in the periphery of the NTD (54). The external regions showing high displacements, namely S71-R78, H146-E156, Q173-Q183 and L249-G257, are all part of an antigenic supersite in the distal-loop region of the NTD (55). In particular, the GTNGTKR motif in S71-R78, besides being an antigenic epitope, has also been suggested to be involved in binding to other receptors, such as sugar receptors (9).

The conformational response of the Q173-Q183 region is of particular interest. The structural response initiated in the FA site propagates through the NTD, reaching the Q173-Q183 segment (Figures 3 and 5). This region, which is located in the distal face of the NTD, forms the entrance of an allosteric site (Figure 5) that has been shown to bind heme (56) and its metabolite biliverdin (57). X-ray crystallography, cryo-EM and mutagenesis experiments together with modelling show that heme and biliverdin bind to a deep cleft in the NTD (56, 57) gated by the Q173-Q183 flexible loop (Figure 5B). Physiological concentrations of biliverdin suppress binding of some neutralising antibodies to the spike (57). Such data suggested a new mode of immune evasion of the spike via the allosteric effect of biliverdin/heme binding (57). In our simulations, two of the three Q173-Q183 regions (in chains A and B) show well-defined outward motions in response to LA removal (Figures 5 and S20). Our results show a clear connection between the FA and biliverdin/heme allosteric sites via internal conformational motions. This connection captured by the D-NEMD approach is remarkable and illustrates the complexity of the potential allosteric modulation of the spike. These results suggest that the presence of heme or its metabolite in the NTD site will affect the internal networks and how dynamic and structural changes are transmitted to and from the FA site. The presence of heme/biliverdin may modulate the response of the spike to fatty acids and vice-versa, potentially affecting the rate and/or affinity of binding of the molecules to their respective allosteric sites. This apparent connection between allosteric sites may be worthy of experimental investigation.

D-NEMD analysis of displacement vectors shows connection between the FA site and the heme/biliverdin binding site in the NTD.

(A) View of the NTD 10 ns after LA removal, focusing on the heme/biliverdin binding site (56, 57) (which is not occupied in the simulations here). Note that the Q173-Q183 segment, which contains residues forming the heme/biliverdin binding site, shows an outward motion upon LA removal. The magnitudes of the displacements are represented on a white-yellow-orange-red colour scale. Vectors with a length ≥0.1 nm are displayed as blue arrows with a scale-up factor of 10. The dark grey spheres represent the FA site. This figure shows the direction of the structural responses around FA site 1 (see Figure S20 for the direction of the motions in the other two FA sites). (B) Cryo-EM structure showing the biliverdin binding site in the NTD (PDB code: 7NT9)(57). The protein is coloured in grey. The biliverdin molecules are shown with spheres.

There are eight N-glycans on the NTD, linked to N17, N61, N74, N122, N149, N165, N234 and N282 (32). From these, four (N74, N122, N149 and N165) are located in or close to the segments that respond to FA site occupancy (Figure S15). N74, N122 and N149 are involved in shielding the spike protein from the host immune system (30), whereas the role of N165 and N234 goes beyond shielding: it is involved in the stabilisation of the RBD ‘up’ conformation (30).

The structural changes induced by LA removal are not restricted to the RBD and NTD: they are transmitted to several regions far from the FA site, notably the furin and S2’ cleavage sites, V620-L629 loop and FP-surrounding regions (Figures 3 and S13-S14). The structural responses starting in the FA site are quickly propagated downwards to the furin cleavage sites and V620-L629 loop via the F318-I326 and C525-K537 segments (Figures 4 and S17-S19). The furin cleavage site, which harbours a polybasic motif containing multiple arginine residues, is located at the boundary between the S1 and S2 subunits and more than 4 nm away from the FA site. This site is important for the protein’s activation, and its removal reduces virus infectivity (26, 50, 52, 58). The addition of extra positively charged residues near the furin cleavage site, as observed in several variants, has been suggested to increase proteolytic processing (59), and has been shown to increase the rate of binding and the affinity of glycosaminoglycans such as heparin and heparan sulfate to this area (60). D-NEMD simulations also show that the addition of extra positive flanking charges, which is observed in some variants (such as P681R in Delta and N679K in Omicron), strengthens the allosteric connection between the FA and furin cleavage site (25). Overall, the allosteric effects are qualitatively similar to what was previously observed for the non-glycosylated spike (25–27): the same regions are connected to the FA site and are significantly affected by ligand removal from this site. This similarity indicates that the presence of glycans on the exterior of the protein does not substantially affect the internal allosteric communication pathways within the spike.

The structural changes starting in the FA site are transmitted to the furin cleavage site and V620-L629, and from there, over time, propagated laterally to the FP proximal region (FPPR) and S2’ cleavage site (Figures 4 and S17-S19). Besides being a known epitope for neutralising antibodies (61, 62), the S2’ cleavage site is also crucial for infection (53, 63). This proteolytic site is located immediately before the hydrophobic FP in the S2 subunit, and its cleavage is mediated by the transmembrane protease serine 2 (TMPRSS2) after binding to the host receptor (63). The FPPR is located after the FP, and it is thought to have a functional role in membrane fusion by mediating the transitions between pre and post-fusion structures of the protein (17). Upon the removal of LA, the residues of the FPPR in direct contact with the FP show a well-defined response upwards towards the C-terminal domain 1 (CTD1) in two of the chains, namely chains B and C, and an outwards motion in chain A. The CTD1 has been suggested (based on cryo-EM structures) to be a structural relay between RBD and FP, sensing the displacement on either side (17)) (Figures 4 and S16-S19). Interestingly, the chains displaying an FPPR motion towards CTD1, notably chains B and C, are the same ones that also exhibit an RBM upward movement towards the solvent. The direction of the S2’ motion observed in the D-NEMD simulations is diverse, with two of the sites (S2’A and S2’B) displaying a motion towards the FP and one showing an opposite movement away from the FP (S2’C) after t = 10 ns (Figures 4 and S16-S19).

The spike contains several complex N- and O-glycans in or close to the furin and S2’ cleavage sites, FPPR, F318-I326, C525-K537 and V620-L629 (32). All three monomers contain one O- and two N-glycans (at positions and T323, N616 and N657) close to the pathway that connects that FA site to the furin cleavage site and FP surrounding regions (Figure S15). Monomer A also contains an additional O-glycan linked to S323 (Figure S15). Notably, interactions between the O-glycans S323 and T325 and the N-glycan at N234 create a direct connection between the NTD and the F318-I326 region of the same monomer (Figure S15). This glycan ‘link’ may facilitate and enhance the transmission of structural changes within an individual monomer. The glycan at position N234 has also been suggested to play a mechanical role in the spike working mechanism by helping to stabilise the RBD in the ‘up’ conformation (30).

To assess whether substitutions, deletions and insertions as seen in the variants lie on the allosteric communication pathways identified here, we overlapped, in the same 3D structure, the spatial positions of the variations with the allosteric networks (Figure S21). In Figure S21, substitutions, deletions and insertions within the allosteric pathways are shown with red spheres, and those close (<0.6 nm away) to the pathways are highlighted in dark blue. The results are interesting: overall, 22 out of the 77 amino acid positions per chain known to vary in the Alpha, Beta, Gamma, Delta and Omicron (BA.1, BA.2, BA.4, BA.5, BQ.1.1 and XBB.1.5) variants directly map onto the allosteric communication pathway identified using the D-NEMD approach. Additionally, a further 28 out of the 77 variations are in direct contact with, or very close proximity to, these networks. H655Y (present in Gamma and all Omicron sub-variants), T547K (in Omicron BA.1), D614G (in Alpha, Betta, Gamma, Delta and all Omicron sub-variants), W856 (in Omicron BA.1) and S982A (in Alpha) are all examples of mutations adjacent to the communication pathways: they modulate the connection to the FA site. These mutations are responsible for the differences in allosteric behaviour observed between SARS-CoV-2 variants. As future variants emerge, it will be of interest to establish if more mutations lie along the allosteric pathways described here. Differences in allosteric behaviour and regulation in the spike are likely to be of functional relevance and useful in understanding differences between SARS-CoV-2 variants.

Conclusions

D-NEMD simulations show important allosteric effects in the fully glycosylated ancestral SARS-CoV-2 spike. We identify the pathways that link the FA site with functional regions for membrane fusion, antigen antibody recognition and allosteric modulation in the fully glycosylated, cleaved spike. These simulations highlight the structural responses resulting from LA removal and demonstrate the complex connection between the FA site and the RBM, an antigenic supersite in the NTD, the allosteric heme/biliverdin binding site, the furin and S2’ cleavage sites and the FP-surrounding regions. The ordering of the conformational changes observed in response to LA removal from the FA sites identifies the initial steps of communication between the different secondary structure elements in the protein, thus constituting the pathways enabling effective signal propagation within the protein. Most tellingly, more than 65% of the substitutions, deletions and additions in the Alpha, Beta, Gamma, Delta, and Omicron variants are located either in or close to the allosteric pathways identified using the D-NEMD approach.

Our findings also show that the presence of glycans on the exterior of the protein, despite affecting the amplitude of the structural responses, does not qualitatively change the cascade of events connecting the FA site to the rest of the spike. Some glycans can influence the allosteric pathways, facilitating the transmission of the structural changes within and between monomers. For example, the interactions between the glycans linked to N234, T373 and S375 can create a direct connection between the NTD and the F318-I326 region of the same monomer, thus helping the propagation of the structural changes within the monomer.

Furthermore, our D-NEMD results reveal an unexpected connection between the FA site and a second allosteric site known to bind heme and its metabolite biliverdin. It will be of interest to understand how heme/biliverdin binding affects the dynamics and structural changes of the spike, and links to the FA site, and potentially other allosteric sites (64). While the effects of the apparent coupling between the heme/biliverdin site and the FA site remains to be investigated, this work has reinforced the ability of the D-NEMD approach to find allosteric sites and to map communication pathways between sites. The results here further point to complex allosteric effects in the SARS-CoV-2 spike, of potential functional relevance.

Supporting information

eLife Assessment

This manuscript focuses on understanding if and how the glycosylation of SARS-CoV2 spike protein affects a putative allosteric network of interactions controlled by the binding of a fatty acid. The main conclusion is that glycans do not significantly affect the network of allosteric interactions. This useful information - albeit mainly consisting of negative results - is based on solid evidence. It will be of interest to scientists focusing on SARS CoV2 protein structure and dynamics.

Reviewer #1 (Public Review):

Summary:

The investigation delves into allosteric modulation within the glycosylated SARS-CoV-2 spike protein, focusing on the fatty acid binding site. This study uncovers intricate networks connecting the fatty acid site to crucial functional regions, potentially paving the way for developing innovative therapeutic strategies.

Strengths:

This article's key strength lies in its rigorous use of dynamic nonequilibrium molecular dynamics (D-NEMD) simulations. This approach provides a dynamic perspective on how the fatty acid binding site influences various functional regions of the spike. A comprehensive understanding of these interactions is crucial in deciphering the virus's behavior and identifying potential targets for therapeutic intervention.

Weaknesses:

The presented evidence is compelling but could be better if this study is supported with sequence analysis to facilitate a complete view of the allosteric networks. The thorough analysis of the simulation results is partially aligned with the discussion because observed in the replicates and the monomers an asymmetry in the perturbations generated by D-NEMD, even when we're using 210 nonequilibrium MD of 10 ns. While the authors claim that the strategy used in this article has been previously validated, the complexity of the spike and the interactions analyzed have required a robust statistical analysis, which is not shown quantitatively. The investigation examines the allosteric modulation within the glycosylated SARS-CoV-2 spike protein, emphasizing the significance of the fatty acid binding site in influencing the structural dynamics and communication pathways essential for viral function, potentially facilitating the development of novel therapeutic strategies. The presented evidence is compelling but needs to be supported by sequence analysis, which will facilitate understanding of the scientific community.

Minor considerations:

Figure S3 shows a discrepancy in the presentation of residue values S325 in the plots of Chains A, B, and C. While chain A shows a value near 0.1, chains B and C plots do not have any value.

Please explain why the plots of figures S6, S7, and S8 show significant changes in several regions, such as RBM and Furin Site. Can these changes be explained?

The flow of the allosteric interaction is complex to visualize just by looking at structures. Could you please include a diagram showing the flow of allosteric interactions (in a sequence diagram or using the structure of the protein)? Or could you include a vector showing how the perturbation done in the FA Active site takes contact with other relevant regions of the Spike protein?

Reviewer #2 (Public Review):

This is a nice paper illustrating the use of equilibrium/non-equilibrium MD simulations to explore allosteric communication in the Spike protein. The results are described in detail and suggest a complex network of signal transmission patterns. The topic is not completely novel as it has been studied before by the same authors and the impact of glycosylation is moderated and localized at the furin site, so not many new conclusions emerge here. It is suggested that mutations are commonly found in the communication pathway which is interesting, but the authors fail to provide evidence that this is related to a positive selection and not simply to a random effect related to mutations at points that are not crucial for stability or function. One interesting point is the connection of the FA site with an additional site binding heme group. It will be interesting to see reversibility, i.e. removal of the ligand at this site is producing perturbation at the FA site?, does it produce other effects suggesting a cascade of allosteric effects? Finally, the paper lacks details to help reproducibility, in particular, I do not see details on D-NEMD calculations. One interesting point is the connection of the FA site with an additional site binding heme group.

Reviewer #3 (Public Review):

Summary:

In a previous study, the authors analyzed the dynamics of the SARS-CoV2 spike protein through lengthy MD simulations and an out-of-equilibrium sampling scheme. They identified an allosteric interaction network linking a lipid-binding site to other structurally important regions of the spike. However, this study was conducted without considering the impact of glycans. It is now known that glycans play a crucial role in modulating spike dynamics. This new manuscript investigates how the presence of glycans affects the allosteric network connecting the lipid binding site to the rest of the spike. The authors conducted atomistic equilibrium and out-of-equilibrium MD simulations and found that while the presence of glycans influences the structural responses, it does not fundamentally alter the connectivity between the fatty acid site and the rest of the spike.

Strengths:

The manuscript's findings are based on an impressive amount of sampling. The methods and results are clearly outlined, and the analysis is conducted meticulously.

Weaknesses:

The study does not clearly show any new findings. The authors themselves acknowledge that the manuscript mainly presents negative results-indicating that glycans do not significantly impact the allosteric network previously reported in other publications. All the results in the paper are based on a single methodology, and additional independent approaches would be needed to confirm the robustness of these findings. Allosteric networks arise from subtle correlations in protein structural dynamics, and it's uncertain whether the results discussed in this manuscript stem exclusively from the chosen force field and other modeling and analysis decisions, or if they indeed reflect something real.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1:

We thank the Reviewer for being very supportive of the work and acknowledging how important it is to understand allosteric modulation in the spike and the potential of this knowledge to contribute to the design of novel therapeutic strategies (for example, disrupting or altering the allosteric networks within the spike can be a novel strategy for drug development against COVID-19). We address their comments below:

(1) The Reviewer states that although the strategy used to extract the responses has been "previously validated", the complexity of the interactions investigated requires "a robust statistical analysis, which is not shown quantitatively".

As the Reviewer points out, the D-NEMD approach has been previously validated in various protein systems ranging from soluble enzymes to integral membrane proteins, including the spike (e.g. [Kamsri et al. (2024) Biochem; Beer et al. (2024) Chem Sci; Oliveira et al. (2023) J Mol Cell Biol; Chan et al. (2023) JACS Au; Castelli et al. (2023) JACS; Castelli et al. (2023) Protein Sci; Oliveira et al. (2022) Comput Struct Biotechnol J; Gupta et al. (2022) Nat Comm; Oliveira et al. (2021) JACS; Galdadas et al. (2021) eLife; Abreu et al. (2019) Proteins; Oliveira et al. (2019) JACS; Oliveira et al. (2019) Structure]. The Kubo-Onsager relation is used to extract the evolution of the protein's response to a perturbation by comparing the equilibrium and nonequilibrium trajectories at equivalent points in time. The calculated responses at individual times are then averaged over all the repeats (210 repeats in the current work), and the standard error of the mean (SEM) is used to assess the significance of the average response. The SEM indicates how much the calculated mean deviates from the true population mean. Calculating the SEM allows us to determine how accurate the measured response is as an estimate of the population response and assess the convergence of our calculations. The evolution of the average Cα displacement and corresponding SEM values for each individual monomer can be visualised in detail in Figures S7-S9. We have added a new sentence to the Materials and Methods section in the Supporting Information, explicitly stating how the convergence and statistical significance of the responses were assessed.

(2) The Reviewer considers that the evidence presented in the paper "is compelling" but suggests performing a sequence analysis to facilitate the understanding of the results by the scientific community.

We thank the Reviewer for their excellent suggestion to perform a sequence analysis of the FA site region and its allosteric connections. Indeed, this analysis (Figure S24) clearly shows that several of the mutations, deletions and insertions in the Alpha, Beta, Gamma, Delta, and Omicron variants are located either in or near the regions of the protein shown to respond to the removal of linoleate from the FA site. These sequence changes affect the protein's responses, and are responsible for the differences in allosteric behaviour observed between variants, as described previously for the non-glycosylated spike [Oliveira et al. (2023) J Mol Cell Biol]. Furthermore, some variants, such as Beta, Gamma, and Omicron, contain residue substitutions at the FA site. For example, the lysine in position 417 in the ancestral spike is mutated to asparagine in Beta and Omicron and threonine in the Gamma variant. Another example is arginine 408 in the original protein, which has been replaced by asparagine in several Omicron sub-variants.

To summarise, the sequence analysis (Figure S24) supports our initial 3D analysis (Figure S25), indicating that many of the changes observed in the variants of concern are indeed in or close to the allosteric networks involving the FA site. We have now included the sequence analysis results in the current paper and added a new figure to Supporting Information showing the sequence alignments between the ancestral spike and different variants (Figure S24).

(3) The Reviewer also has "minor considerations": first, they point to a discrepancy in the presentation of residue values S325 in the plots of Chains A, B, and C of Figure S3; second, they ask why several regions, such as RBM and Furin Site in figures S6, S7, and S8 show significant changes.

To answer both points raised by the Reviewer, we need to start by explaining that the spike typically features 22 N-glycosylation and at least two O-glycans sites per monomer. These sites have been found to be heterogeneously populated in different experimental studies (e.g. [Watanabe et al. (2020) Science; Shajahan et al. (2020) Glycobiology; Zhang et al. (2021) Mol Cell Proteomics]). Given this, the spike model used as the starting point for this work reflects this heterogeneity, with asymmetric site-specific glycosylation profiles derived from the glycoanalytic data reported by Watanable et al. for N-glycans [Watanabe et al. (2020) Science] and Shajahan et al. for O-glycans [Shajahan et al. (2020) Glycobiology]. This means that the glycan occupancy and composition for each site differ between the three monomers. For example, while monomer A contains the two O-glycans sites (linked to T323 and S325, respectively) fully occupied, monomers B and C only contain the T323 O-glycan. A detailed description of the glycosylation of the spike model is given in the supporting information of [Casalino et al. (2020) ACS Cent Sci].

Regarding the Reviewer's first minor point, the discrepancy in behaviour observed in Figure S3 for S325 is related to the fact that this glycosylation site is only occupied in monomer A, with no glycans present in this site in monomers B and C.

Regarding the second point, the differences observed in the responses between the three monomers in Figures S7-S9 are probably due to asymmetries in the protein dynamics introduced by the different glycosylation patterns in the monomers.

We have now added a new paragraph to the materials and methods section in the Supporting Information describing the asymmetric site-specific glycosylation profiles of the monomers.

(4) Due to the complexity of the allosteric interactions observed, the Reviewer suggests including in the paper a "diagram showing the flow of allosteric interactions" or a "vector showing how the perturbation done in the FA Active site takes contact with other relevant regions".

This is an excellent suggestion to facilitate the visualisation of the allosteric networks. We have added a new figure to Supporting Information highlighting the allosteric pathways identified from the DNEMD simulations and the direction of the propagation of the structural changes (Figure S26).

Reviewer #2:

We thank the Reviewer for their time in evaluating our manuscript and providing suggestions for improving it and ideas for further work. We are happy that the Reviewer found this to be a "nice paper" with the calculations "well done" and interesting results. We address their comments below:

(1) The Reviewer suggests improving the paper by adding a more detailed explanation of the DNEMD simulations approach, a method that, although proposed decades ago, is still generally unfamiliar to the community. They also asked for "information on the convergence of the observables".

As stated by the Reviewer, a dynamical approach to nonequilibrium molecular dynamics (D-NEMD) was first proposed in the seventies by Ciccotti et al. [Ciccotti et al. (1975) Phys Rev Lett; Ciccotti et al. (1979) J Stat Phys]. This approach combines MD simulations in equilibrium and nonequilibrium conditions. The rationale for the D-NEMD approach is simple and can be described as follows: if an external perturbation (e.g. binding/unbinding of a ligand) is added to a simulation sampling an equilibrium state and, by doing so, a parallel nonequilibrium simulation is started, the structural response of the protein to the perturbation can be directly measured by comparing the equilibrium and nonequilibrium trajectories at equivalent points in time by using the Kubo-Onsager relation as long as enough sapling is gathered (for more details, please see the reviews [Balega et al. (2024) Mol Phys; Oliveira et al. (2021) Eur Phys J B; Ciccotti et al. (2016) Mol Simul]). This approach, although conceptually simple, is very powerful as it allows for computing the evolution of the dynamic response of the protein to the external perturbation, while assessing the convergence and statistical significance of that response. This approach also has the advantage that the convergence and significance of the response can be easily evaluated, and the associated errors can be computed and made as small as desirable by increasing the number of nonequilibrium trajectories. Determining the statistical errors associated with the responses (through, e.g., the determination of the standard error of the mean, SEM) is essential to test if the sampling gathered is sufficient. In this paper, the SEM was calculated for each average Cα displacement value at times 0.1, 1 and 10 ns after the removal of linoleate, LA (see Figures S7-S9). The SEM indicates how accurate the measured response is as an estimate of the population response and allows us to assess the convergence of the results.

Generally, multiple (tens to hundreds) D-NEMD simulations are needed to achieve statistically significant results for biomolecular systems (for examples, see [Balega et al. (2024) Mol Phys; Oliveira et al. (2021) Eur Phys J B]). As such, the length of the D-NEMD simulations (typically 5 to 10 ns) reflects the balance between the computational resources available and the number of replicates needed to achieve statistically significant responses from the system. Following the Reviewer's suggestion, we have now added a brief description of the D-NEMD approach to the main manuscript and expanded the D-NEMD section in the Supporting Information with a more detailed description of the method, including adding a new figure showing a schematic representation of the D-NEMD approach (Figure S5) as well as explicitly stating the settings used in these simulations and how the statistical significance of the responses was assessed.

(2) The Reviewer suggests comparing the D-NEMD results with "more traditional analysis, such as correlation analysis, or community network analysis".

We agree with the Reviewer that this is an important comparison, which can provide a broader, more articulate and coherent picture of spike allostery and have, therefore, performed additional analysis. The dynamic cross-correlation analysis suggested by the Reviewer is a valuable tool for identifying the regions in the protein influenced by the FA site in equilibrium conditions. However, such an approach is not straightforwardly applicable to D-NEMD simulations, as these simulations are not in equilibrium. Nevertheless, as suggested by the Reviewer, we have determined the cross-correlation matrices for both the equilibrium and D-NEMD simulations (Figure S22), similar to those in our previous work [Galdadas et al. (2021) eLife] and [Oliveira et al. (2022) J Mol Cell Biol]. The analysis of these matrices can provide information about possible allosteric networks. In Figure S22, the cyan and blue regions represent moderate and high negative correlations between Cα atoms, while orange and red regions correspond to moderate and high positive correlations. Negative correlations indicate residues moving in opposite directions (moving toward or away from each other). In contrast, positive values imply that the residues are moving in similar directions. We also note that, with collaborators, we have compared D-NEMD and other nonequilibrium and equilibrium MD analysis methods for allostery [Castelli et al. (2023) JACS].

The cross-correlation maps depicted in Figure S22 show moderate to high positive correlations between the FA sites and two of the three RBDs in the protein. This happens because each FA site sits at the interface between two neighbouring RBDs. Low to moderate negative and mildly positive correlated motions can also be observed between the FA site and the NTDs and fusion peptide surrounding regions, respectively. To facilitate the visualisation of the above-described motions, we have also mapped the statistical correlations for R408 and K417 (two FA site residues able to directly form salt-bridge interactions with the carboxylate head group of LA) on the protein's three-dimensional structure (Figure S23). Figure S23 highlights the patterns of movement described above and allows us to identify the regions whose motions are coupled to the FA site.

Interestingly, some segments forming the signal propagation pathways, such as R454-K458 in all three monomers, and C525-K537 in monomers B and C, can also be identified from the cross-correlation matrices, showing moderate to high correlations with the FA site (Figures S22-S23). The crosscorrelation maps computed from the equilibrium trajectories (with FA sites occupied with LA) show a slight increase in the dynamic correlations, mainly for the RBDs, compared to the maps obtained from the nonequilibrium trajectories (Figure S22). This indicates that the presence of LA in the FA strengthens the connections between the FA site and other parts of the protein.

We have updated the manuscript to include the cross-correlation analysis, with two new figures added to Supporting Information: one depicting the cross-correlation maps for the D-NEMD and equilibrium simulations (Figure S22), and the other showing the statistical correlations for R408 and K417 (Figure S23).

(3) The Reviewer considers the observed connection between the fatty acid site and the heme/biliverdin site "interesting" and suggests "exploring the impact of ligand removal on this secondary site on the protein".

Similarly to the Reviewer, we find the connection between the FA and the heme/biliverdin site fascinating and worthy of further investigation. The observed connection between these two sites shows the complexity of the allosteric effects in the spike. It would be interesting and informative to perform new equilibrium simulations of the heme/biliverdin spike complex and a new set of D-NEMD simulations in which this site is perturbed (e.g. through the removal of the heme group) to map the networks connecting this allosteric site to other functionally important regions of the spike, including the FA site and potentially other allosteric sites. These new simulations would allow us to assess the reversibility of the connection between the FA and heme/biliverdin sites and enhance our understanding of allosteric modulation in the spike and the role of the heme/biliverdin site in this process. However, due to the large size of the system and the associated computational demands, such simulations are not possible within the timeframe of the revision of this paper. These simulations would take many months to complete using our HPC resources. We also note that an experimental structure of the spike containing both heme and linoleate is not available. Further simulation analysis of the communication pathways involving the heme/biliverdin site is an excellent idea for future work.

(4) The Reviewer "liked the mapping of existing mutations on the communication pathway" and suggested a more detailed study focusing on the effect of the mutations.

We fully agree with the Reviewer and consider that a detailed study focusing on the effect of the mutations, insertions, and deletions in the different glycosylated variants of concern (including new emerging ones) would be of great interest. Our previous work using D-NEMD on the non-glycosylated ancestral, Alpha, Delta, Delta plus and Omicron BA.1 spikes revealed significant differences in the allosteric responses to LA removal, with the changes in the variants affecting both the amplitude of the structural responses and the rates at which these rearrangements propagate within the protein [Oliveira et al. (2023) J Mol Cell Biol].

Using the D-NEMD approach to systematically investigate the impact of each individual mutation and their contribution to the overall allosteric response of the glycosylated variants (similar to what we have done previously for the D614G mutation in the non-glycosylated protein [Oliveira et al. (2021) Comput Struct Biotechnol J]) would provide insights into the functional modulation of the spike. However, as noted above in point 3, spike simulations are highly computationally expensive, both in terms of processing and data storage requirements, because of the large size of the protein and the need for equilibrium and D-NEMD simulations. This makes the suggested mutational study unfeasible within the timeframe of the current revisions. It is, however, an excellent idea for future research.

Reviewer #3:

We thank the Reviewer for carefully reading and critically reviewing this work and recognising that the findings reported are "based on an impressive amount of sampling" and "meticulous" analysis. We address their comments below:

(1) The Reviewer considers that this work "does not clearly show any new findings" as it shows that the glycans do not significantly impact the internal networks in the protein.

We respectfully disagree with the Reviewer. This work identifies new allosteric effects in the spike, specifically, the connection of the FA site with the heme binding site. The equilibrium simulations alone provide the first analysis of the effects of linoleate binding in the fully glycosylated spike. The finding that glycosylation does not significantly affect the allosteric pathways in the spike is in itself an important finding. Previous D-NEMD simulations investigated only the non-glycosylated spike ([Oliveira et al. (2021) Comput Struct Biotechnol J; Oliveira et al. (2022) J Mol Cell Biol] ) leading to questions of whether the allosteric effects pathways were changed by glycosylation; our results here show that the main conclusions are reinforced, but glycosylation does have some effect on networks, and also on the speed of the dynamical response. To the best of our knowledge, our work represents the first investigation to analyse the impact of glycosylation on the allosteric networks in the spike. We show that even though the presence of glycans in the exterior of the spike does not significantly alter the internal communication pathways in the protein, in some cases (for example, the glycans linked to N234, T373 and S375), they create direct connections between different regions, which may facilitate the propagation of the structural changes.

(2) The Reviewer suggests adding a "clear and concise description" of the D-NEMD approach to the manuscript.

We appreciate that the use of the D-NEMD method to study biomolecular systems is relatively new, and so may be unfamiliar. As explained above in our response to Reviewer 2 (point 1), a brief description of the D-NEMD approach was now included in the main manuscript. A detailed description of the method was also added to Supporting Information, including a new figure representing the rationale for the approach (Figure S5). The interested reader is directed to previous applications and reviews for more details of the method (e.g. [Balega et al. (2024) Mol Phys; Oliveira et al. (2021) Eur Phys J B; Ciccotti et al. (2016) Mol Simul; Kamsri et al. (2024) Biochem; Beer et al. (2024) Chem Sci; Oliveira et al. (2023) J Mol Cell Biol; Chan et al. (2023) JACS Au; Castelli et al. (2023) JACS; Castelli et al. (2023) Protein Sci; Oliveira et al. (2022) Comput Struct Biotechnol J; Gupta et al. (2022) Nat Comm; Oliveira et al. (2021) JACS; Galdadas et al. (2021) eLife; Abreu et al. (2019) Proteins; Oliveira et al. (2019) JACS; Oliveira et al. (2019) Structure]).

(3) The Reviewer invites us to "discuss the robustness of the findings with respect to forcefield choices".

The Reviewer raises an important but rather complex question, and one which can, of course, be posed for any molecular dynamics simulation study. The short answer is that we have chosen state-of-the-art forcefields, which have been shown to give results for the spike that are in good agreement with experiments; glycosylated spike simulations are rather computationally expensive, and constructing the models also requires significant human time and effort. Thus, while in principle interesting, it is not practical to repeat the current simulations with different forcefields. However, as detailed below, comparison of our simulations of the glycosylated and non-glycosylated [Oliveira et al. (2022) Comput Struct Biotechnol J] spike using different forcefields indicates that our conclusions are robust and are not dependent on the choice of forcefield.

Comparing the performance and accuracy of different force fields is not straightforward, as the results depend on the system of interest, properties simulated and sampling. In this work, the CHARMM36m all-atom additive force field was used to describe the protein and glycans. CHARMM36m is a widely used force field that has previously been validated for the simulations of biological systems [Huang et al. (2013) J Comput Chem; Guvench et al. (2009) J Chem Theory Comput], including proteins, lipids and glycans, with many of studies adopting it in the literature. Additionally, the glycosylated models of the spike used in this work have also been successfully applied and tested before (e.g. [Dommer et al. (2023) Int J High Perform Comput Appl; Sztain et al. (2021) Nat Chem; Casalino et al. (2021) Int J High Perform Comput Appl; Casalino et al. (2020) ACS Cent Sci]), with their dynamics shown to correlate well with experimental data.

It is also worth pointing out that, despite differences in the amplitude of the responses, the allosteric networks identified using the D-NEMD approach for the non-glycosylated [Oliveira et al. (2022) Comput Struct Biotechnol J] and glycosylated spikes are generally similar (Figure S13). While the responses for the non-glycosylated protein were extracted from simulations using the AMBER99SBILDN forcefield [Oliveira et al. (2022) Comput Struct Biotechnol J], those reported in this work were obtained from trajectories using the CHARMM36m forcefield. The similarity between the responses for the two systems (which were simulated using different forcefields) is a good indication that our findings are forcefield independent.

(4) The Reviewer suggests comparing our findings with "alternative methods of analysing allostery".

As stated above in our response to Reviewer 2 point 2, we consider the suggested comparison an excellent idea. We have therefore performed a dynamic cross-correlation analysis to identify the regions in the protein coupled to the FA site in both equilibrium and nonequilibrium conditions (see Figures S22-S23). Overall, this analysis shows that the FA site motions are strongly coupled to the RBDs and moderately to weakly connected to the NTDs and fusion peptide surrounding regions (please see a detailed description of the results of the correlation analysis in our response to Reviewer 2 point 2). The cross-correlation analysis performed was added to the manuscript, and two new figures were included in the Supporting Information (Figures S22-S23): the first, showing the cross-correlation maps for the D-NEMD and equilibrium simulations; the second, showing the statistical correlations for R408 and K417 (two residues forming the FA site and that can directly interact with the carboxylate head group of LA).

We agree that comparing different allosteric analysis methods is interesting, informative and important. As noted above, we have compared D-NEMD and other nonequilibrium and equilibrium MD analysis methods for allostery in the well-characterised K-Ras system [Castelli et al. (2023) JACS].

Version 2

Abstract

Abstract

The trimeric spike protein plays an essential role in the SARS-CoV-2 virus lifecycle, facilitating virus entry through binding to the cellular receptor angiotensin-converting enzyme 2 (ACE2) and mediating viral-host membrane fusion. The SARS-CoV-2 spike contains a fatty acid (FA) binding site at the interface between two neighbouring receptor-binding domains. This site, also found in some other coronaviruses, binds free fatty acids such as linoleic acid. Binding at this site locks the spike in a non-infectious, closed conformation. This site is coupled to functionally important regions, but the effects of glycans on these allosteric effects have not been investigated. Understanding allostery and how this site modulates the behaviour of the spike protein could potentiate the development of promising alternative strategies for new coronavirus therapies. Here, we apply dynamical nonequilibrium molecular dynamics (D-NEMD) simulations to investigate allosteric effects of the FA site in the fully glycosylated spike of the original SARS-CoV-2 ancestral variant. The results show allosteric networks that connect the FA site to important functional regions of the protein, including some more than 40 Å away, including the receptor binding motif, an antigenic supersite in the N-terminal domain, the furin cleavage site, regions surrounding the fusion peptide, and another allosteric site known to bind heme and biliverdin. The networks identified here highlight the complexity of the allosteric modulation in this protein and reveal a striking and unexpected connection between different allosteric sites. Notably, 65% of amino acid substitutions, deletions and insertions in the Alpha, Beta, Delta, Gamma and Omicron variants map onto or close to the identified allosteric pathways. Comparison of the FA site connections from D-NEMD in the glycosylated and non-glycosylated spikes revealed that the presence of glycans does not qualitatively change the internal allosteric pathways within the protein, with some glycans facilitating the transmission of the structural changes within and between subunits.

Main text

The SARS-CoV-2 virus, like other β-coronaviruses, uses the spike protein to mediate virus entry into host cells. The spike is a trimeric glycoprotein embedded in the virus envelope. During the initial stage of the SARS-CoV-2 infection process, the spike binds to host target cells, primarily via the receptor angiotensin-converting enzyme 2 (ACE2) (1, 2), but it can also bind to other targets, such as neuropilin-1 (3, 4), estrogen receptor α (5) and potentially to nicotinic acetylcholine receptors (6–9) and sugar receptors (10). Given its crucial role in the infection process and the fact that it is one of the main targets for antibody neutralization, the spike is one of the most important targets for developing COVID-19 therapies and vaccines (e.g. (11–15)).

Each spike monomer comprises three regions: a large ectodomain, a transmembrane domain and a short cytoplasmic tail (16–18). The ectodomain, the main focus of this work, is composed of two subunits (S1 and S2) and contains the structural motifs that directly bind to the host receptors as well as those needed for the membrane fusion process (16–18). S1 is responsible for binding to the human ACE2 receptors, while S2 for the fusion of the viral and host membranes (16–18). The spike also contains three equivalent fatty acid (FA) binding sites at the interfaces between neighbouring receptor binding domains (RBDs) (19) (Figure 1A). Each FA binding site is a hydrophobic pocket formed by two RBDs, with one RBD providing the aromatic and hydrophobic residues to accommodate the FA hydrocarbon tail and the other providing the polar and positively charged residues that bind the FA carboxylate headgroup (Figure 1B). The essential FA linoleic acid binds (as linoleate, LA) with high affinity to the FA pocket, stabilizing the spike in a non-infectious locked conformation, in which the RBDs are all ‘down’ with the receptor-binding motifs (RBMs) occluded inside the trimer, and thus inaccessible for binding to ACE2 (19). The discovery of this site inspired the development of new spike-based potential therapies based on FAs or other natural, repurposed, or specifically designed small molecules able to bind to the FA site (11, 20–24). Several cryo-EM structures of the SARS-CoV-2 spike in complex with small molecules, such as linoleic, oleic, and all-trans retinoic acid and SPC-14, bound to the FA site, are now available (11, 19–21). Following this discovery, equivalent FA sites have been identified in several closely related coronavirus spikes (19, 25, 26). Surface plasmon resonance experiments and cryo-EM structures show that the FA site is conserved in the spike proteins of highly several pathogenic β-coronaviruses, such as SARS-CoV, MERS-CoV, SARS-CoV-2, but not in the spikes of common, mild disease-causing β-coronaviruses (27).

Structure of the glycosylated head region of the ancestral SARS-CoV-2 spike with linoleate bound to the free fatty acid (FA) binding site.

(A) Model of the ectodomain of the glycosylated SARS-CoV-2 spike with linoleate (LA) bound. The spike-LA complex model was built using the cryo-EM structure 7JJI as a reference (26). Each monomer in the spike homotrimer is shown in a different colour: dark blue, light blue and green. Glycans are indicated with grey sticks, and LA molecules are highlighted with magenta spheres. Three FA binding sites exist in the trimer, each located at the interface between two neighbouring monomers. In this model, all three receptor-binding motifs (RBMs) are in the ‘down’ conformation, and the protein is cleaved at the furin recognition site at the S1/S2 interface. (B) Detailed view of the FA binding site. This hydrophobic site is formed by two RBDs, with one providing the hydrophobic pocket for the FA hydrocarbon tail and the other providing polar (Q409) and positively charged (R408 and K417) residues to bind the negatively charged FA headgroup.

Simulations using the dynamical-nonequilibrium molecular dynamics (D-NEMD) approach showed that the FA site allosterically modulates the behaviour of functional motifs in both the ancestral (also known as wild type, ‘early 2020’, or original) spike and in several variants (28–30), and that these effects differ between variants. These D-NEMD simulations (which tested the effects of removing linoleate) showed that the FA site is allosterically connected to the RBM, N-terminal domain (NTD), furin cleavage site, and the region surrounding the fusion peptide (FP) (28–30). These regions have significantly different allosteric behaviours between the ancestral, Alpha, Delta, Delta plus, and Omicron BA.1 variants (28–30). They differ not only in the amplitude of the structural responses of these regions but also in the rates at which the structural changes propagate (28–30). However, these previous D-NEMD simulations did not incorporate the spike’s many N- and O-linked glycans, which are critical to its function, not only in protecting it from immune recognition, but also in modulating its dynamics (31–34). A crucial unresolved question, therefore, is whether and how glycosylation affects allosteric communication with the FA site.

The ancestral spike is heavily glycosylated with 22 predicted N-linked glycosylation sites per monomer (16, 35), of which 17 have been found to be occupied (35–38). The ancestral spike also contains at least two O-glycosylation sites per monomer with low occupancy (37, 39, 40). The occupancy and composition profile of the glycosylation sites differ between variants, expression systems, and experimental methods (41, 42). This glycan coating plays a crucial role in shielding the virus from the immune system (33, 34), and in infection (31–34, 43). Glycans also affect the dynamics and stability of essential regions of the protein, including the RBDs, and modulate binding to ACE2 (31–33). Here, we use D-NEMD simulations (44–46) to characterize the response of the fully glycosylated SARS-CoV-2 ancestral spike to LA removal, and investigate allosteric modulation by the FA site and the effects of glycans on the protein’s allosteric behaviour. In recent years, D-NEMD simulations (44–46) have emerged as a powerful computational approach to investigate a diversity of biological problems from the trasmission of structural changes (47, 48) to the identification of allosteric effects (28–30, 49–55) and the impact of pH changes (56) in fundamentally different biomolecular systems, including SARS-CoV-2 targets (28–30, 54, 56). For example, in the SARS-CoV-2 main protease, dynamical responses from D-NEMD pinpointed positions associated with drug resistance (54), and for the spike, D-NEMD indicated the regions of the protein affected by pH changes (56).

We performed extensive equilibrium MD simulations, followed by hundreds of D-NEMD simulations, to analyze the response of the fully glycosylated, cleaved (at the furin recognition site) spike to LA (linoleate) removal. Principal component analysis was performed to check the equilibration and sampling of the equilibrium replicates (Figure S1C). In the equilibrium simulations, the locked state of the spike (with all RBDs down) with linoleate bound remained stable (Figure S1A), showing structural convergence after ∼50 ns and minimal secondary structure loss after 750 ns (Figure S1B). The comparison between the average Cα fluctuations calculated from the equilibrium simulations for the locked (this work) and closed (from Casalino et al. (33)) glycosylated spikes shows that the dynamics of the protein is generally similar (Figure S1D). The largest difference is observed in RBMB, which exhibits decreased dynamics in the locked state (i.e. when LA is present in the FA sites) (Figure S1D). In the locked spike equilibrium simulations, all LA molecules remained stably bound to the protein (Figures S2A and S2D), with the carboxylate head group of LA making consistent salt-bridge interactions with K417 and occasional interactions with R408 (Figures S2B and S2C).

Analysis of the dynamics of the glycans in the equilibrium trajectories (Figures S3-S4) showed (as previously observed for the spike without LA (33)) that the glycans are very mobile, exhibiting diverse levels of motion depending on their composition, branching and solvent exposure (Figure S3). Generally, N-glycans in the NTD show higher fluctuations than those of the RBD (Figure S3). The glycan linked to N331 from chain C is an exception, with one of the largest RMSF values. The O-glycans connected to T323 and S325, close to the RBD, are less flexible than the N-glycans (Figure S3). The highly dynamic profile of the glycans (Figure S4) helps the spike to evade the host immune response by masking immunogenic epitopes, thus preventing them from being targeted by the host’s neutralizing antibodies. To quantify the shielding effect, the spike accessible surface area (ASA) covered by the glycans was determined for probe radii ranging from 0.14 nm (approximate radius of a water molecule) to 1.1 nm (approximate radius of a small antibody molecule). As can be seen in Figures S4B and S4C, consistently with previous findings reported for the closed state (with all RBDs in the ‘down’ conformation, without LA bound) (33), in the locked state, the spike head has a thick glycan shield, which covers ∼60% of the protein accessible area for a 1.0-nm-radius probe and restricts the binding of medium size molecules to the protein. However, small molecules (probes with a radius 0.14-0.3nm) can penetrate the shield more easily as it only covers ∼26% of the area of the protein accessible to smaller probes (Figure S4C).

An ensemble of 210 conformations (70 configurations per replicate) was extracted from the equilibrium MD simulations and used as starting points for the D-NEMD simulations, which investigated the effect of LA removal (Figure S6). The D-NEMD method, originally proposed by Ciccotti et al (57, 58), combines simulations in equilibrium and nonequilibrium conditions (Figure S5). It allows for the direct computing of the evolution of the dynamical response of a system to an external perturbation. The rationale for the D-NEMD can be described as follows: if an external perturbation is applied to an equilibrium simulation and, by doing so, a parallel nonequilibrium simulation is started, then the response of the protein to the perturbation can be straightforwardly extracted using the Kubo-Onsager relation (for more details, see (44, 46)).

The perturbation used here, the instantaneous removal of LA from all three FA sites, is the same as in previous D-NEMD simulations of non-glycosylated spikes (28–30). This perturbation takes the system out of equilibrium, and creates a driving force for changes to occur as the protein responds to the perturbation and relaxes back towards equilibrium. LA removal from the FA sites triggers the structural response of the protein as it adapts to an empty FA site. Analysis of the evolution of the structural changes reveals the mechanical and dynamical coupling between the structural elements involved in response to LA removal and identifies the allosteric pathways connecting the FA site to the rest of the protein. The evolution of the structural response of the protein is extracted using the Kubo-Onsager relation (44, 45, 57, 59) from the difference between the equilibrium and nonequilibrium trajectories at equivalent points in time (Figure S7). The response obtained for each pair of equilibrium and nonequilibrium simulations is averaged over all 210 trajectories here, hence reducing noise (44, 45), and allowing the statistical significance of the responses to be assessed from the standard error of the mean (Figure S7-S9) (44).

LA removal initiates a complex chain of structural changes that are, over time, propagated within the protein. The deletion of the LA molecules immediately triggers a structural change in the FA site, which contracts as the sidechains of the residues lining it move closer to each other, filling the space once occupied by the LA molecule (Figure S10). The changes in the FA site are then swiftly transmitted to well-defined regions of the protein, notably the NTD, RBM and FP-surrounding regions (Figures 2 and S11-S12).

Structural response of the glycosylated spike to LA removal.

The average Cα displacements 0.1, 1 and 10 ns after LA removal from the FA binding sites are shown, mapped onto the starting structure for the equilibrium simulations. The norm of the average Cα displacement vector between the D-NEMD apo and equilibrium LA-bound simulations was calculated for each residue using the Kubo-Onsager relation (44, 45, 57, 59). The final displacement values are the averages obtained over the 210 pairs of simulations (Figures S7-S9). The cartoon thickness and structure colours (scale on the right) indicate the average Cα-positional displacement. Each RBD, NTD, furin site and FP are subscripted with their chain ID (A, B or C). Glycans are shown as light grey sticks, whereas the dark grey spheres highlight the position of the LA molecule. The FA site shown in this figure is FA site 1, which is located at the interface between chains C and A (see Figures S11-S12 for the protein responses from the viewpoint of FA site 2 and 3, respectively). The responses of all three FA sites are qualitatively similar, with the same motifs and sequence of events observed.

The cascade of events observed here for the fully glycosylated ancestral spike mirrors that of the non-glycosylated protein (28–30), with LA removal triggering immediate structural changes in the FA site, which are then transmitted to key regions of the protein, including to the FP-surrounding region which is more than >40 Å away from the FA site. Figure S13 shows a strong positive correlation between the responses obtained for the non-glycosylated and glycosylated spikes, underscoring the overall similarity between our current and previous D-NEMD simulations. The largest differences in the responses between the glycosylated and non-glycosylated spikes are found around the furin recognition site (Figure S13). This is unsurprising as the furin site is cleaved in the current (glycosylated) simulations but remains uncleaved in the previous non-glycosylated simulations. The furin cleavage/recognition site is a polybasic four-residue insertion located on a solvent-exposed loop at the S1/S2 junction (16, 17). This site is important for the activation of the spike (60), and its presence affects viral infectivity (e.g. (29, 60–63)).

The evolution of the response of the spike to LA removal reveals the pathways through which structural changes propagate from the FA site to functional motifs (e.g. motifs involved in membrane fusion and antigenic epitopes (Figures 3 and S14-S15)). These pathways lie mainly within the protein and are generally similar to those previously found for the non-glycosylated spike (28–30). The structural changes induced by LA removal, which start in the FA site (mainly in the P337-A348 and S366-A372 regions of one monomer and T415-K417 of the other one), are rapidly transmitted to the rest of the RBD. The R454-K458 region is particularly important as it mediates the transmission of the structural changes to the A475-C488 segment in the RBM (Figure 3A).

Structural responses of functional regions of the spike.

Close-up view of the structural response of the RBD (A), NTD (B) and FP surrounding regions (C) to LA removal. The FA site shown here is FA site 1, located at the interface between chains C and A (see Supporting Figures S14-S15 for the responses of the other two FA sites, which are similar). Structure colours and cartoon thickness indicate the average Cα displacement values. Each region is subscripted with its chain ID (A, B or C). The dark grey spheres show the FA binding site. In the images representing the spike at t=0 ns (left side images A, B and C), the glycans are shown as light grey sticks. Glycans were omitted from the figures showing the responses at t=0.1, 1 and 10 ns to facilitate visualization. For more detail, see the legend of Figure 2.

In addition to the amplitude of the structural changes induced by LA removal, the average directions of the motion can also be computed by determining the average displacement vector of Cα atoms (54) between the equilibrium and nonequilibrium trajectories at equivalent time points (Figures 4 and S17-S20). Indeed, the amplitude of the structural changes in Figures 2-3, S11-S12 and S14-S15, corresponds to the norm of the average Cα displacement vector. Upon LA removal, the spike regions that form the FA site show responses with well-defined directions. These segments include P337-A348 and S366-A372, which are regions with residues whose side chains form the FA pocket (Figures 4 and S17-S20). Soon after LA deletion, P337-A348 and S366-A372 (P337A-A348A and S366A-A372A in FA site 1, P337B-A348B and S366B-A372B in FA site 2 and P337C-A348C and S366C-A372C in FA site 3) move inwards towards the FA site. These motions collectively reflect the contraction of the FA site upon LA removal. The directions of RBM motions are more diverse, with two of the RBMs, namely RBMC (Figure S18) and RBMB (Figure S19), displaying an upward movement and the third showing an opposite downward one (RBMA) at t=10 ns (Figures 4 and S20).

Directions of the structural responses of the RBD and FP-surrounding regions to LA removal.

The average Cα displacement vectors at t=10 ns are shown. These vectors were determined by averaging Cα displacement vectors between the equilibrium and nonequilibrium trajectories over the 210 replicas. Vectors with a length ≥0.1 nm are displayed as blue arrows with a scale-up factor of 10. The average displacement magnitudes are represented on a white-yellow-orange-red scale. The dark grey spheres represent the FA site. This figure shows the directions of the responses around FA site 1, which is located at the interface between chains C and A (see Supporting Figures S19-S20 for the direction of the motions around the other two FA sites).

The ancestral SARS-CoV-2 spike has two glycans located on the RBD: N-glycans at positions N331 and N343 (35). The N343 glycan is particularly interesting because it is situated immediately after one of the regions that transmits structural changes from the FA site, namely P337-F342 (Figure 3). As well as being close to the FA site, this glycan can also directly bridge the two neighbouring RBDs (Figure S16), which may strengthen the connection between the FA site and these regions. N343 has been shown to play a role in the RBD opening mechanism by acting as a gate for the change from the ‘down’ to the ‘up’ conformation (32).

The structural changes induced by LA removal are also swiftly propagated to the NTD via P337-F342, W353-I358, and C161-P172 (Figures 3 and S14-S15). Despite variations in amplitude, the structural response to LA removal is consistent across the three FA sites, exhibiting the same motifs and sequence of events for signal propagation to the NTDs. The response of the protein, which starts in the P337-A348 segment in the FA site, is transmitted to W353-I358, C161-P172 and then to several antigenic epitopes located in the periphery of the NTDs (64). The external regions showing high displacements, namely S71-R78, H146-E156, Q173-Q183 and L249-G257, are all part of an antigenic supersite in the distal-loop region of the NTD (65). In particular, the GTNGTKR motif in S71-R78, besides being an antigenic epitope, has also been suggested to be involved in binding to other receptors, such as sugar receptors (10).

The conformational response of the Q173-Q183 is of particular interest as this region forms a pocket that binds heme and the tetrapyrrole products of its metabolism (66, 67). The structural response initiated in the FA site propagates through the NTD, reaching the Q173-Q183 segment (Figures 3, 5 and S14-S15). This region, which is located in the distal face of the NTD, forms the entrance of an allosteric site (Figure 5) that has been shown to bind heme (66) and its metabolites biliverdin and bilirubin (67). X-ray crystallography, cryo-EM and mutagenesis experiments together with modelling show that heme and biliverdin bind to a deep cleft in the NTD (66, 67) gated by the Q173-Q183 flexible loop (Figure 5B). Physiological concentrations of biliverdin suppress binding of some neutralizing antibodies to the spike (67). Such data suggested a new mode of immune evasion of the spike via the allosteric effect of biliverdin/heme binding (67). In our D-NEMD simulations, two of the three Q173-Q183 regions (in chains A and B) show well-defined outward motions in response to LA removal (Figures 5 and S20). Our results show a clear connection between the FA and biliverdin/heme allosteric sites via internal conformational motions despite the two sites being >30 Å apart. This connection, captured by the D-NEMD approach, is remarkable and illustrates the complexity of the potential allosteric modulation of the spike. These results suggest that the presence of heme or its metabolite in the NTD site affects the internal networks and how dynamic and structural changes are transmitted to and from the FA site. The presence of heme/biliverdin may modulate the response of the spike to fatty acids and vice-versa, potentially affecting the rate and/or affinity of binding of the molecules to their respective allosteric sites. This apparent connection between allosteric sites is worthy of further investigation.

D-NEMD displacement vectors show a connection between the FA site and the heme/biliverdin binding site in the NTD.

(A) View of the NTD 10 ns after LA removal, focusing on the heme/biliverdin binding site (66, 67) (which is not occupied in the simulations here). Note that the Q173-Q183 segment, which contains residues forming the heme/biliverdin binding site, shows an outward motion upon LA removal. The magnitudes of the displacements are represented on a white-yellow-orange-red colour scale. Vectors with a length ≥0.1 nm are displayed as blue arrows with a scale-up factor of 10. The dark grey spheres represent the FA site. This figure shows the direction of the structural responses around FA site 1 (see Figure S21 for the direction of the motions in the other two FA sites). (B) Cryo-EM structure showing the heme/biliverdin binding site in the NTD (PDB code: 7NT9)(67). The protein is coloured in grey. The biliverdin molecules are shown with spheres.

There are eight N-glycans on the NTD, linked to N17, N61, N74, N122, N149, N165, N234 and N282 (35). Of these, four (N74, N122, N149 and N165) are located in or close to the segments that respond to FA site occupancy (Figure S16). N74, N122 and N149 are involved in shielding the spike protein from the host immune system (33), whereas the role of N165 and N234 goes beyond shielding: they are involved in the stabilization of the RBD ‘up’ conformation (33).

The structural changes induced by LA removal are not restricted to the RBD and NTD: they are also transmitted to several regions far from the FA site, notably the furin site at the S1/S2 boundary, the S2’ protease recognition and cleavage site, R815, located the S2 subunit (immediately upstream the fusion peptide), V620-L629 loop and FP proximal region, FPPR (Figures 3 and S14-S15). The structural responses starting in the FA site quickly propagate downwards to the furin cleavage sites and V620-L629 loop via the F318-I326 and C525-K537 segments (Figures 4 and S18-S20). The furin cleavage site, which harbours a polybasic motif containing multiple arginine residues, is located at the boundary between the S1 and S2 subunits and more than 40 Å from the FA site. The furin cleavage site is important for protein activation, and its removal reduces viral infectivity (29, 60, 62, 68). The addition of extra positively charged residues near the furin cleavage site, as observed in several variants, has been suggested to increase proteolytic processing (69), and has been shown to increase the rate of binding and the affinity of glycosaminoglycans such as heparin and heparan sulfate for this area (70). In our previous work, the comparison of the D-NEMD responses to LA removal between variants of concern has also suggested that the addition of extra positive flanking charges, which is observed in some variants (such as P681R in Delta and N679K in Omicron), strengthens the allosteric connection between the FA and furin cleavage site (28). Overall, the allosteric effects observed here for the glycosylated ancestral spike are qualitatively similar to those in the non-glycosylated protein: the same regions are connected to the FA site and are affected significantly by ligand removal from this site (28–30). This indicates that the glycans on the exterior of the protein do not substantially affect the internal allosteric communication pathways within the spike.

The structural changes starting in the FA site are transmitted to the furin cleavage site and V620-L629, and from there, over time, propagated to the FPPR and S2’ cleavage site (Figures 4 and S18-S20). In addition to being an epitope for neutralizing antibodies (71, 72), the S2’ cleavage site is also crucial for infection (63, 73). This proteolytic site is located immediately before the hydrophobic FP in the S2 subunit, and its cleavage is mediated by the transmembrane protease serine 2 (TMPRSS2) after binding to the host receptor (73). The FPPR is located after the FP, and it is thought to have a functional role in membrane fusion by mediating the transitions between pre- and post-fusion structures of the protein (18). Upon removal of LA, the residues of the FPPR in direct contact with the FP show a well-defined response upwards in two of the chains, namely chains B and C, and an outwards motion in chain A (Figures 4 and S19-S20). FPPRB and FPPRC, located closer to FA sites 1 and 2 respectively, show an upward movement towards the C-terminal domain 1 (CTD1), whereas FPPRA, which is nearer to FA site 3, exhibits a lateral outward motion. The CTD1 has been suggested (based on cryo-EM structures) to be a structural relay between RBD and FP, sensing displacement on either side (18)). Interestingly, the chains displaying FPPR motion towards CTD1, notably chains B and C, also exhibit an RBM upward movement away from the body of the spike. The direction of the S2’ motion observed in the D-NEMD simulations is diverse, with two of the sites (S2’A and S2’B) displaying a motion towards the FP and one showing an opposite movement away from the FP (S2’C) after t = 10 ns (Figures 4 and S17-S20).

The spike contains several complex N- and O-glycans in or close to the furin and S2’ cleavage sites, FPPR, F318-I326, C525-K537 and V620-L629 (35). All three monomers contain one O- and two N-glycans (at positions T323, N616 and N657) close to the pathway that connects that FA site to the furin cleavage site and FP regions (Figure S16). Monomer A also contains an additional O-glycan linked to S323 (Figure S16). Notably, interactions between the O-glycans S323 and T325 and the N-glycan at N234 create a direct connection between the NTD and the F318-I326 region of the same monomer (Figure S16). This glycan ‘link’ may facilitate and enhance the transmission of structural changes within an individual subunit. The glycan at position N234 has also been suggested to play a mechanical role in the spike infection mechanism by helping to stabilize the RBD in the ‘up’ conformation (33).

Cross-correlation analysis was performed for the equilibrium and D-NEMD simulations (similarly to references (28, 55)) to identify the coupled regions in the protein, including those with motions connected to the FA site. In Figure S22, the dark and light blue regions represent high and moderate negative correlations between the Cα atoms in the protein, and red and orange regions correspond to high and moderate positive correlations. Negative correlation values indicate that the atoms are moving in opposite directions, whereas atoms systematically moving along the same direction show strong positive correlations. Overall, the cross-correlation maps computed from the equilibrium and D-NEMD trajectories show similar patterns, with the former exhibiting a subtle increase in the correlations between the FA sites and RBDs (Figure S22). This increase indicates that binding a fatty acid molecule, such as LA, to the FA site reinforces the connection between this site and other parts of the protein.

The cross-correlation maps also show positive correlations between each FA site and two of the three RBDs in the protein. This is because each FA site sits at the interfaces between every two neighbouring RBDs (e.g. FA site 1 is formed by residues from subunits A and C). Low to moderate negative and small positive correlated motions are observed between the FA site and the NTDs and fusion peptide regions, respectively (Figures S22). To visualize these motions, the statistical correlations for R408 and K417 (two FA site residues able to form salt-bridge interactions with the carboxylate head group of LA) were mapped on the protein structure (Figure S23). Figure S23 shows the patterns of movement described above and the regions whose motions are connected to the FA site. Interestingly, some segments forming the signal propagation pathways, such as R454-K458 in all three monomers and C525-K537 in monomers B and C can also be identified from the cross-correlation analysis, showing moderate to high correlations with the FA site (Figures S22-S23).

To assess whether the substitutions, deletions and insertions seen in various SARS-CoV-2 variants lie on the allosteric communication pathways identified, we overlapped, in the same 3D structure, the sequence variations (using spheres to highlight the position of the changes) with the D-NEMD responses reported above (Figure S25). In Figure S25, changes within the allosteric pathways are indicated by red spheres, while those within 0.6 nm of any atom (i.e., both main and sidechain atoms) forming the paths are highlighted in dark blue. The results are interesting: 22 of the 77 amino acid positions per chain that differ in the Alpha, Beta, Gamma, Delta and Omicron (BA.1, BA.2, BA.4, BA.5, BQ.1.1 and XBB.1.5) variants directly map onto the allosteric communication pathway identified using D-NEMD. A further 28 out of the 77 variations are in direct contact with, or very close proximity to, these networks. H655Y (present in Gamma and all Omicron sub-variants), T547K (in Omicron BA.1), D614G (in Alpha, Betta, Gamma, Delta and all Omicron sub-variants), W856K (in Omicron BA.1) and S982A (in Alpha) are all examples of mutations close to the communication pathways, which may influence the connection to the FA site. These mutations are responsible for the previously observed differences in allosteric behaviour between SARS-CoV-2 variants (28).

Sequence alignment of the original spike and the Alpha, Beta, Gamma, Delta and Omicron (BA.1, BA.2, BA.4, BA.5, BQ.1.1 and XBB.1.5) variants (Figure S24) shows that several of the mutations, deletions and insertions present in the variants are located either in or near the pathways identified here. Furthermore, some variants, such as Beta, Gamma, and Omicron, contain residue substitutions at the FA site. For example, the lysine in position 417 in the original spike is mutated to asparagine in Beta and Omicron and threonine in the Gamma variant. Another example is arginine 408 in the Ancestral protein, which has been replaced by asparagine in several Omicron sub-variants. As future variants emerge, it will be of interest to establish whether mutations lie in the FA and heme/biliverdin (66, 67) sites or along the allosteric pathways described here. Differences in allosteric behaviour and regulation in the spike are likely to be of functional relevance and may be useful in understanding differences between SARS-CoV-2 variants. Our previous work using D-NEMD revealed significant differences in the allosteric responses of spike variants to LA removal (28). The substitutions, insertions, and deletions in the variants affected both the amplitude of the structural responses and the rates at which these rearrangements propagate within the protein (28). The allosteric connections identified in Alpha were generally similar to the ancestral protein, whereas Delta exhibited increased connections between the FA site and the furin cleavage site but diminished links to V622–L629 (28). Omicron displayed significant changes in the NTD, RBM, and furin cleavage site, with stronger couplings observed between the FA site and these regions (28).

Conclusions

D-NEMD simulations show important allosteric effects in the fully glycosylated ancestral SARS-CoV-2 spike. These simulations identify the pathways that link the FA site with functional regions (regions of the spike involved in membrane fusion, antibody recognition and allosteric modulation). The D-NEMD simulations show the structural responses resulting from LA removal and demonstrate connections between the FA site and: the RBM; an antigenic supersite in the NTD; the allosteric heme/biliverdin binding site (66, 67); the furin site; and, the FP-surrounding region (including S2’ cleavage site and, the FPPR) (Figure S26). The connection between the FA site and the RBM is mediated by residues R454-K458, while the transmission of structural changes from the FA site to the NTD involves residues P337-F342, W353-I358, and C161-P172. The allosteric pathways connecting the FA site to the furin site and FP-surrounding region involve the segments containing residues F318-I326 and C525-K537. Notably, more than 65% of the substitutions, deletions and additions in the Alpha, Beta, Gamma, Delta, and Omicron variants are located either in or close to the allosteric pathways identified using D-NEMD.

Furthermore, our D-NEMD results reveal an unexpected connection between the FA site and a second allosteric site known to bind heme and its metabolites (66, 67). It will be of interest to understand how heme/biliverdin binding affects the dynamics and structural changes of the spike, and links to the FA site, and potentially other allosteric sites (74). While the effects of the apparent coupling between the heme/biliverdin site and the FA site remain to be investigated, this work has reinforced the ability of the D-NEMD approach to find allosteric sites and to map communication pathways between sites (44, 46). The results here further point to the complex allosteric effects in the SARS-CoV-2 spike, of potential functional relevance.

Comparison with previous D-NEMD simulations of the non-glycosylated spike (28–30) shows that the presence of glycans on the exterior of the protein, does not qualitatively change the cascade of events connecting the FA site to the rest of the spike. Some glycans influence the allosteric pathways, facilitating the transmission of the structural changes within and between subunits. For example, the interactions between the glycans linked to N234, T373 and S375 can create a direct connection between the NTD and the F318-I326 region of the same monomer, thus helping the propagation of structural changes within the monomer. These results shed new light on the roles of glycans and further emphasise their potential in modulating the functional dynamics of the spike.